---
title: "temp"
format:
  pdf:
    documentclass: scrartcl
    fontsize: 11pt
    geometry:
      - top=1in
      - bottom=1in
      - left=1.25in
      - right=1.25in
    number-sections: false
    colorlinks: true
editor: visual
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

knitr::opts_chunk$set(echo = FALSE)
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 4)

library(nbastatR)
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(stringi)
library(ggcorrplot)
library(glmnet)
library(lme4)
library(caret)

# Last complete season
season = 2024
player_df = read_csv("player_23-24.csv")
all_nba = read_csv("all-pro_23-24.csv")

# View the advanced statsa
advanced_stats = bref_players_stats(seasons = season, tables = "advanced") |>
  select(namePlayer, ratioWS, ratioOWS, ratioDWS, ratioWSPer48, ratioOBPM, ratioDBPM, ratioBPM, ratioVORP) |>
  arrange(desc(ratioWS))

# Clear large df out of memory
dataBREFPlayerAdvanced = NULL

all_nba = all_nba |>
  pivot_longer(
    cols = c(`1st Team`, `2nd Team`, `3rd Team`),
    names_to = "all_nba",
    values_to = "name"
  ) |>
  mutate(name = str_trim(name))

player_df = player_df |>
  left_join(all_nba, by = join_by(Player==name)) |>
  mutate(Player = stri_trans_general(Player, "latin-ascii"),
         is_all_nba = as.numeric(!is.na(all_nba))) |>
  # All-NBA qualification criteria
  filter(G>=65,
         MP>=65*20-10) |>
  left_join(advanced_stats, by = join_by(Player==namePlayer))

```

## STA540 Case Study 2 Draft

------------------------------------------------------------------------

Project Title: All-NBA Selection Prediction

Data citation: Sports Reference LLC "2024-25 NBA Player Stats: Totals" Basketball-Reference.com - Basketball Statistics and History. <https://www.basketball-reference.com/>

Original data collection date:

Data pull date: 02/03/2026

Acronyms:

-   NBA - National Basketball Association
-   GLMM - generalized mixed effect model
-   AUC - Area under the curve
-   PR - precision-recall
-   ROC - receiver operating characteristic
-   ratioWS - win share

------------------------------------------------------------------------

### 1 Introduction

#### Background/Introduction:

The All-NBA Team is one of the National Basketball Association’s most prestigious individual honors. It is awarded to the players that have performed at the highest level during the regular season. The selections have been determined through voting by a panel of 100 sportswriters and broadcasters, with players ranked based on ballot points. The top five players are assigned to first (best), the next five best players to the second team, and 11th-15th ranked players to the third team. These awards not only count towards a player's legacy, but also their salary (contract incentives), reputation, and Hall of Fame consideration.

Advances in computing and game-tracking have led to new basketball metrics and available data. Advanced metrics like win share attempt to quantify player contribution to a team's wins in one number to objectively measure player performance. Modelling all-NBA selections using objective metrics not only proves or disproves the value of some of these metrics, but can also highlight the impact subjective measures like voter fatigue, reluctance to repeatedly vote for the same dominant players, or popularity have on NBA awards.

#### Study Aims

The study will examine player box scores at the end of a season and calculate advanced statistics, namely win share, to identify if All-NBA status can be predicted from metrics using a generalized linear mixed effect model (GLMM). The primary aim of the study will be to determine the predictive value of win shares on all-NBA selection. Secondary aims will look at other advanced metrics like box-plus-minus in comparison with win share.

### 2 Data

The data for this study comes from Basketball Reference, a website that aggregates several basketball statistics. The key variables of interest are `is_all_nba`, a binary variable indicating whether a player was selected, and `ratioWS`, which is a rational number representing the win share of the player. Win share is a metric that attempts to divide up a team's wins to its player proportionally by contribution (https://www.basketball-reference.com/about/ws.html).

The dataset includes NBA players with 65 games played and average over 20 minutes in those games, based on selection criteria for the award. There is no missing data, due to meticulous stat-tracking.

### 3 Exploratory Data Analysis

In @fig-WS-All-NBA, there is a clear distinction in win shares across the 4 different selection groups. This indicates that win shares can be used as a useful predictor. There is, however, some overlap between the all-NBA groups and some of the unselected players, meaning that win shares alone cannot predict all-NBA status.

```{r}
#| label: fig-WS-All-NBA
#| fig-width: 5
#| fig-height: 3

ggplot(player_df, aes(x = factor(all_nba), y = ratioWS, fill = factor(all_nba))) +
  geom_boxplot(outlier.color = "red") +
  scale_x_discrete(labels = c("0" = "Not Selected", "1" = "All-NBA")) +
  labs(title = "Win Shares: All-NBA vs. Non-All-NBA Players",
       subtitle = "WS is expected to be significantly higher for selected players",
       x = "All-NBA Selection", y = "Total Win Shares") +
  theme_light() +
  theme(legend.position = "none")
```

### 4 Model Fitting and Diagnostics

#### Statistical Model

The probability of selection is modeled using a logistic link:

$$
\text{logit}(p_i) =
\beta_0 +
\beta_1 \text{ratioWS}_i +
\beta_2 G_i +
\beta_3 MP_i +
\beta_4 FG_i +
\beta_5 FGA_i +
\beta_6 FT_i +
\beta_7 FTA_i +
\beta_8 TRB_i +
\beta_9 AST_i +
\beta_{10} STL_i +
\beta_{11} BLK_i +
\beta_{12} TOV_i +
\beta_{13} PF_i +
\beta_{14} PTS_i
$$

where all predictor variables are standardized.

To perform shrinkage and automatic variable selection, the regression coefficients are assigned a horseshoe prior:

$$
\beta_j \ | \ \tau^2, \lambda_j^2 \sim \mathcal{N}(0, \tau^2 \lambda_j^2), \quad j = 1, \ldots, 14
$$

$$
\lambda_j \sim C^{+}(0,1)
$$

$$
\tau \sim C^{+}(0,\tau_0)
$$

where $C^{+}(0,1)$ denotes the half-Cauchy distribution. The global parameter $\tau$ controls overall shrinkage, while the local parameters $\lambda_j$ allow individual coefficients to escape shrinkage when strongly supported by the data.Let $Y_i$ denote whether player $i$ was selected to an All-NBA team. $\tau_0$ indicates the shrinkage corresponding to the expected number of true predictors (7 out of 14 in this case) (Carvalho 2009).

#### Model Selection and Diagnostics

Because the initial frequentist model did not converge, I fit a Bayesian logistic regression with a horseshoe prior to shrink estimates that are multicollinear with a small dataset. Due to the relatively small dataset available, methods that rely on hyperparameter tuning like LASSO or Ridge regression are unsuitable for variable selection. All predictors were standardized, so each coefficient corresponds to a one standard deviation increase in the predictor. Posterior sampling mixed well (all $\hat{R}\approx 1.00$ and effective sample sizes $>1000$), indicating convergence and stable posterior estimates.

### 5 Discussion

The posterior suggests that win share is the strongest predictor of All-NBA selection. Not only does it have the largest coefficient by magnitude, but it is also the only strongly positive coefficient, with an credible interval that does not contain 0 ($\beta=2.91$, 95% CrI: (0.75 to 6.19). This supports the primary aim that win shares contain meaningful a predictive signal for All-NBA outcomes. Other variables such as minutes played (MP), steals (STL), blocks (BLK), and personal fouls (PF) had posterior intervals overlapping zero in this fit, suggesting weaker predictive value.

Since there are only two seasons since the all-NBA eligibility rules changed, the 2023-2024 season data was used to fit the model validated against the 2024-2025 season outcomes. The model succesfully identified 12 out of the 15 (~80%) all-NBA selections. 

#### Limitations

The validation and training sets are both quite small, containing only 15 positive values each. The small datasets means that the choice of prior is very important in determining the posterior. The model used weakly informative prior on scaled predictors, where the choice of expected predictors was arbitrary. This means that there is room for improved prior selection. 

In addition, the model also ignores the time series structure of all-NBA voting. For instance, a player's previous season's performance might somehow affect voter's perception of the player's current season performance. Any correlation across time is ignored. 

#### Conclusion

This case study demonstrates that objective metrics can predict All-NBA selection with meaningful accuracy, with **PTS** and **ratioWS** emerging as the most informative predictors among those considered. However, the modeling goal is fundamentally constrained by the award structure (only 15 selections) and by unmeasured factors such as media narratives, player popularity, and contextual team performance. Future work could incorporate team record, betting odds/award narratives, and position/role indicators, and could evaluate performance using cross-validation and precision-recall metrics tailored to class imbalance.