---
title: "Case Study 2 Project Proposal"
author: "Luxman Maheswaran"
format: pdf
editor: visual
---

#### Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
```

**Project choice \# (1-5 based on Canvas descriptions): #4 Sports Reference**

**Dataset(s) within selected source, if applicable: Basketball**

**External data, if applicable:**

**Brief description: Player Stats by Season & All Pro Selections**

The primary dataset would be a season-long aggregation of a player's box score. Advanced metrics like [win-shares](https://www.basketball-reference.com/about/ws.html) can be calculated from these statistics. In addition, there is a separate data set that contains All Pro selections for a season.

The other research question would use season-long team statistics (same as player statistics aggregated on a team level).

**Research question 1: Can box scores and advanced metrics successfully predict All Pro selections using a generalized linear mixed effects model?**

-   Outcome variable (include the name/description and type of variable): All Pro Selection (Binary - Yes/No)

**Research question 2 (optional): Can team statistics predict playoff teams, and if so how far in advance?**

-   Outcome variable (include the name/description and type of variable): Playoff Entrance (Binary - Yes/No)

**Load the data and provide a `glimpse()`**:

```{r data-1, warning=FALSE, message=FALSE}

library(tidyverse)
library(stringr)

player = read_csv("player_24-25.csv") |>
  mutate(Player = str_squish(Player))

all_pro = read_csv("all-pro_24-25.csv") |>
  pivot_longer(
    cols = everything(),
    names_to = "All-Pro",
    values_to = "Player") |>
  mutate(Player = str_squish(Player))

player = player |>
  left_join(all_pro, by = join_by(Player == Player))

glimpse(player)
```
